The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-09-17 paper.qmd
made some edits based on Susans comments

Uncertainty visualisation is a relatively new field in research. 
Early papers that specifically reference "uncertainty visualisation" appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display [@MacEachren1992; @Carr1992]. 
@fig-ibrekk depicts an example of the uncertainty visualisations discussed in these early papers. 
Despite the new terminology visualisation of uncertainty has been present since the earliest times. 
For example, box plots or histograms can be considered to be displaying uncertainty in the sense of variability in observations sampled from a population distribution. 
Today, there are an abundance of publications on the topic which makes it timely to construct a review of the field. 
In fact, there have already been several reviews published.

<!-- Susan: I think there's a missed opportunity here to reference the premise you started with, that when there are heaps of information, it's valuable to distill it down into simple facts -->

@@

# Visualising uncertainty as a variable
Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious. 
You may think "well, I will just add a dimension to my plot that includes uncertainty". 
This is a reasonable approach. 
The simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel. 
However, it is unclear if this approach is 
sufficient XXX Word choice??? XXX
<!-- I changed will to sufficient, but I'm not sure that's what you were going for. -->
for our purposes.

@@

Descriptive statistics describe our sample as it is and summarizes large data down into an easy to swallow format. 
Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case. 
In 19th century England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson). 
Practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations [@Otsuka2023]. 
In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature", that is, we need to assume that unobserved phenomena should be similar to observed phenomena [@Otsuka2023]. 
Positivists abhor the assumption of the "uniformity of nature" as they believed referencing the unobservable is bad science. 
In other words, these scientists embraces descriptive statistics and shunned inferential statistics due to the inherent uncertainty that came with them. 
Since uncertainty is non-existent in descriptive statistics, it is clear that uncertainty is a by-product of inference.
<!-- Inferential statistics didn't yet exist when positivists were dominant - they weren't making a choice between description and inference, as the choice didn't exist. This is slightly misleading as written, even if I generally agree with the concept you're trying to communicate. -->

>>>>

Uncertainty visualisation is a relatively new field in research. 
Early papers that specifically reference "uncertainty visualisation" appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display [@MacEachren1992; @Carr1992]. 
@fig-ibrekk depicts an example of the uncertainty visualisations discussed in these early papers. 
Despite the new terminology visualisation of uncertainty has been present since the earliest times. 
For example, box plots or histograms can be considered to be displaying uncertainty in the sense of variability in observations sampled from a population distribution. 
Today, there are an abundance of publications on the topic which makes it timely to construct a review of the field.
That is, now that there is an overwhelming amount of information, it is valuable to distill it into simple facts. 
In fact, there have already been several reviews published.

@@

# Visualising uncertainty as a variable
Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious. 
You may think "well, I will just add a dimension to my plot that includes uncertainty". 
This is a reasonable approach. 
The simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel. 
However, it is unclear if this approach is sufficient for our purposes.

@@

Descriptive statistics describe our sample as it is and summarizes large data down into an easy to swallow format. 
Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case. 
In 19th century England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson). 
Practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations [@Otsuka2023]. 
In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature", that is, we need to assume that unobserved phenomena should be similar to observed phenomena [@Otsuka2023]. 
Positivists believed referencing the unobservable was bad science. 
In other words, these scientists embraced descriptive statistics due to the inherent certainty that came with them.
Since uncertainty is non-existent in descriptive statistics, it is clear that uncertainty is a by-product of inference.
